Gravitational Lens Recovery with \Glass: How to measure the mass profile and shape of a lens

### 2013-07-10 limits/ms.tex
editing introduction

@@

\section{Introduction}\label{sec:intro} %-----------------------------------------------------
Since the discovery of the first gravitational lens
\citep{1979Natur.279..381W}, lensing has become an increasingly important probe
of the distribution of mass in the Universe
\citep{1937ApJ....86..217Z,2012arXiv1206.1225A}. Strong lensing in particular 
provides a powerful probe of the mass distribution at the heart of galaxies and
galaxy clusters \citep{2010CQGra..27w3001B,2007ApJ...667..645R,2006ApJ...652L...5S,2009ApJ...690..154S}, giving valuable constraints on the nature of dark matter \citep{2006ApJ...652L...5S,2013ApJ...765...25N}, cosmology 
\citep{2010Sci...329..924J, 2008ApJ...679...17C}, and galaxy formation \citep{2012MNRAS.424..104L}. To date, some $\sim$400 strong lenses are
known\footnote{{\tt http://admin.masterlens.org/}.}, but with the advent of
large ground and space based surveys, many thousands of lenses are expected to
be discovered over the next ten years \citep{2012arXiv1206.1225A,
2010AAS...21540115M, 2004NewAR..48.1085K}.

The majority of lens modelling efforts to date have focussed on `parametric'
modelling, where simple theoretically motived models are fit using maximum
likelihood methods to the data \citep{2011A&ARv..19...47K, 1993A&A...273..367K,
2010GReGr..42.2151K}. This is useful in that a simple model provides valuable
physical insight. However, as data quality improves and the questions we ask of
these data become more refined, it is important to re-asses the impact of our
model assumptions. `Non-parameteric' methods -- that assume many more
parameters than the available data constraints -- are invaluable in this
respect \citep{1997MNRAS.292..148S,2005MNRAS.360..477D,2010ApJ...723.1678C,2006MNRAS.367.1209L,2013arXiv1304.2393S}. Here, the lens inversion problem is
deliberately chosen to be under-constrained.
Instead of finding one `best-fit' model, we must search for many models
consistent with the data, exploring degeneracies between different but
plausible\footnotemark\ mass models
\citep{2000AJ....120.1654S,2006ApJ...653..936S,2008MNRAS.386..307L}. The key
advantages of non-parametric modelling are that we can: (i) explore model
degeneracies and marginalise over them; and (ii) we can determine what type and
quality of data are required to measure parameters of interest.

\footnotetext{Constraints extra to the available data are typically required on the solution space in order to make it finite. Such constraints -- priors -- are typically chosen to eliminate theoretically implausible solutions (an extreme example might be if all of the mass lies in one corner of the lens model space, or if the mass fluctuates wildly on the smallest scales probed). Such priors are typically far less restrictive than the implicit constraints inherent in parametric mass modelling.}

>>>>

\def\Sigmacrit{\Sigma_{\rm crit}}

@@

\section{Introduction}\label{sec:intro} %-----------------------------------------------------

Strong gravitational lenses are rare.  The reason is that the line of
sight to a light source in the universe rarely goes through a region
such that the projected density in its sky neighbourhood exceeds the
critical density to produce multiple images.  To date, some $\sim400$
strong lenses are known,\footnote{See {\tt http://www.masterlens.org}
  for a catalogue.}  but this number is expected to increase to
several thousand over the next ten years.  The optimism derives from
new and upcoming surveys, both
ground-based\footnote{http://pan-starrs.ifa.hawaii.edu}$^{\rm,}$\footnote{http://www.darkenergysurvey.org}
and spacecraft,\footnote{http://www.euclid-ec.org} together with a
community of citizen-science volunteers examining the image data for
candidates.\footnote{http://spacewarps.org}

Since lensing depends only on gravity, strong lenses are a rather
unique window on dark matter, and indirectly on the cosmological model
as well, and hence the thousands of to-be-discovered lenses are
eagerly awaited \citep{2010CQGra..27w3001B,2012arXiv1206.1225A}.
Extracting results from lensing, however, will require modelling, and
it is towards that goal that the present work is directed.

To see why lens modelling is of crucial importance, let us recall the
essential quantities that appear in lensing.  First we have the
distances: let $D_L$, $D_S$, $D_{LS}$ be the angular-diameter
distances to the lens, source, and from lens to source; there are all
strictly $\propto c/H_0$.  Typically
\begin{equation}
D_L \approx z \frac c{H_0}\ \hbox{and}\ \frac{D_S}{D_{LS}} \sim 1.
\end{equation}
The critical density for multiple images is
\begin{equation}
\Sigmacrit = \frac{c^2 D_L}{4\pi G} \frac{D_S}{D_{LS}}
\end{equation}
in mass per steradian.  The separation between the lensed images is of
order the Einstein radius $\theta_E$.  The image separation is related
to the mass by
\begin{equation}
\theta_E \sim \frac{R_G}{D_L} \frac{D_{LS}}{D_S}
\end{equation}
where $R_G$ is the gravitational radius $GM/c^2$.  If the source is a
quasar or otherwise rapidly variable, a time delay is the variability
will be present
\begin{equation}
\Delta t \sim R_G/c \,.
\end{equation}
So in principle, one can not only measure the mass of the lens, one
can use the dependence on the cosmology-dependent $D$ factors to
extract the cosmological model and all its parameters.
\cite{1937ApJ....86..217Z} drew attention to the former, and
\cite{1964MNRAS.128..307R,1966MNRAS.132..101R} pointed out the latter,
all long before lenses were discovered.  The difficulty became
apparent soon after the discovery of the first lens by
\cite{1979Natur.279..381W}.  In the first ever paper on lens
modelling, \cite{1981ApJ...244..736Y} found that many plausible mass
distributions could reproduce the data.

% Simple parametric models are usually robust on qualitative features,
% but other aspects can be very model-dependent
% \citep[cf.~][]{2003AJ....125.2769S}.  For example, for lensed
% quasars, there may be several models all providing excellent fits to
% the image data but disagreeing on the expected time delays, but they
% almost always agree about the {\em order\/} of time delays. Thus, a
% simple model provides valuable physical insight, but as data quality
% improves and the questions we ask of these data become more refined,
% it is important to re-assess the impact of our model assumptions.
% The key advantages of non-parametric modelling are that we can: (i)
% explore model degeneracies and marginalise over them; and (ii) we
% can determine what type and quality of data are required to measure
% parameters of interest.

\cite{1981ApJ...244..736Y} were remarkably prescient about the
subsequent development of lens modelling.  First, the introduced the
technique of choosing a parametric form for the lensing mass and then
fitting for the parameters, which is still the most common strategy
\citep[see for example][]{2010GReGr..42.2151K,2011A&ARv..19...47K}.
Second, they pointed out the non-uniqueness of lens models, or lensing
degeneracies.  Third, they suggested combining lensing data with
stellar kinematics and X-rays, to reduce the effect of the
degeneracies.  Later work, as well as following up these suggestions,
has introduced some ideas.  Four of these are important in the present
work.
\begin{itemize}
\item In free-form models, there is no specified parametric form for
  the mass distribution.  There are still assumptions (or priors) on
  the mass distribution, such as smoothness or being centrally
  concentrated
  \citep{1997MNRAS.292..148S,2005MNRAS.360..477D,2010ApJ...723.1678C}
  but these are much less restrictive than parametric forms.
  \cite{2006MNRAS.367.1209L} implemented a particular elegant prior,
  which is that the mass distribution must be non-negative and no
  extra images are produced.  Free-form are at present most common for
  cluster lenses\citep{2009ApJ...690..154S,2013arXiv1304.2393S}.
\item Model ensembles, exploring a diverse range of possible mass
  distributions that nonetheless all fit the data, are a way of
  cambating the non-uniqueness of models.  Model ensembles are
  possible in parametric models \citep{1999AJ....118...14B} but are
  more common in free-form models
  \citep{2000AJ....119..439W,2009ApJ...690..154S,2012MNRAS.425.3077L}.
\item Stellar kinematics has played a role in many modelling
  techniques.  The underlying reason is the virial theorem, which
  relates the line of sight velocity dispersion to the Einstein radius
  as
\begin{equation}
  \frac{\langle v^2_{\rm los}\rangle}{c^2} \approx
  \frac{\theta_E}{4\pi} \frac{D_S}{D_{LS}}
\end{equation}
  with the relation becoming exact for isothermal lenses.  The use of
  two-dimensional kinematics by \cite{2011MNRAS.415.2215B} is
  especially interesting.
\item The stellar mass in a lens can be inferred from photometry and
  compared with the total mass
  \citep{2005ApJ...623L...5F,2008MNRAS.383..857F,2011ApJ...740...97L}.
  Since the inferred stellar mass does depend on the assumed IMF,
  lenses in which stellar mass dominates can be used to derive upper
  bounds on the stellar $M/L$ \cite{2010MNRAS.409L..30F}. Lower bounds
  on stellar $M/L$ have also been claimed \citep{2013MNRAS.428.3183D}
  but these require upper-bound assumptions about the dark-matter
  profile.
\item Testing modelling strategies to see how well they recover
  simulated lenses is increasingly recognized as essential.  Simple
  blind tests have appeared in earlier work \citep[for example, Figure
    2 in][]{2000AJ....119..439W}, but more recently, test against
  dynamically simulated galaxies are favoured
  \citep{2007ApJ...667..645R,2009MNRAS.393.1114B}.
\end{itemize}
There are also more modelling ideas in the literature.  One is to use
X-ray intensity and temperature profiles as a mass constraint
\citep[such as in][]{2013ApJ...765...25N}.  Another idea is to model
multiple lenses simultaneously, with one or more cosmological
parameters variable but shared between the lenses.  This strategy has
been used to constrain $H_0$ from time delays
\citep{2006ApJ...652L...5S,2008ApJ...679...17C,2010ApJ...712.1378P}.
In principle, cosmological parameters could be constrained even from a
single lens.  Some models have attempted this, starting from a prior
based on the values and uncertainties from CMB fluctuations
\citep{2010Sci...329..924J,2013arXiv1306.4732S}, but tests using
simulated lenses are desirable.
